Toward theories of deliberated preferences

### 2020-10-24 simple.tex
Mostly review, improved (unfinished)

\section{Species of theories}
\label{sec:approaches}
For this article, a \emph{normative decision theory} is a theory that, when applied in the situations it self-describes as applicable to, claims to yield valid prescriptions. I consider a prescription in the wide sense: recommending against some behavior, or recommending conditionally, also counts as prescribing.

A \emph{descriptive decision theory} is a theory that, when correctly applied in the situations it self-describes as applicable to, claims to predict the decision behavior of a given set of individuals. For this article, a claim counts as a prediction only when it is in principle possible to show that it is false using empirical data, thus predictions only apply to empirically observable data.

@@

These definitions avoid to define the concept of theory itself, for defining precisely such a fundamental concept raises serious difficulties. Is Prospect theory in general a theory, or only some of its variants, sufficiently precisely defined? Can a theory be extremely specific, or is some level of generality required to deserve being called a theory? Must a theory consist in a non-trivial assembly of laws, some being deducible from others? It is not necessary for the purpose of the present discussion to take position on these issues, rather, the focus is on the distinction between normative and descriptive approaches. An intuitive understanding of the concept of theory will suffice.
Similarly, I sidestep insuperable difficulties pertaining to capturing in a short definition the concept of a descriptive theory in general: some descriptive theories may propose to describe without predicting, for example (historical decisions or trends come to mind); such endeavor simply does not enter the scope of this discussion, and I make no claim about its legitimacy to be called a descriptive theory in general.

I say that a normative theory is also a descriptive theory when the prescriptions of the normative theory, in supplement of claiming to be valid prescriptions (in the situations that the theory claim to be applicable to), also claim to predict the decision behavior of a given set of individuals.

Another contrast can now be introduced: the contrast between shallow and reflexive behavior. \emph{Shallow behavior} is defined as the everyday, usual behavior of an individual. 
This is the behavior whose observation defines what \citet[p.\ 16]{von_neumann_theory_2004} describe as the “intuitive” preferences of an individual, those originating from an “immediate sensation” (cited in \citet{fishburn_retrospective_1989}).
\emph{Reflexive behavior}, by contrast, is defined as the behavior that enacts informed, pondered decisions, those that would be taken after due consideration.
% an individual has consciously adopted after having pondered everything that is possibly relevant to the decision. 
%This pair of definitions is, again, vague and unsatisfactory for a deep philosophical discussion, but hopefully sufficient for the present one. 
%What is meant by “pondering about everything that is possibly relevant to the decision” will be defined more precisely later; an intuitive understanding of this phrase will do at this stage.
This distinction draws on our capacity, as human beings, to think about our shallow preferences and consider some of them as mistaken \citep{frankfurt_freedom_1971}. 
A famous example of such a situation is the one of \citet[pp.\ 101--103]{savage_foundations_1972} having realized, after due consideration, that his initial declared preference in a decision problem presented by \citet{allais_so-called_1979} was mistaken.

This contrast raises a new distinction, which appears by simple combination of the just given definitions: a \emph{shallow descriptive theory} is a descriptive theory that claims to predict the shallow behavior of individuals; a \emph{reflexive descriptive theory} is one that claims to predict the reflexive behavior of individuals.

The descriptive theories proposed in decision theory, in the sense defined here, are exclusively shallow descriptive theories, to the best of my knowledge. (\Cref{sec:laundering} discusses the theories that launder preferences.)
%This need not be so. Indeed, this article introduces descriptive theories about reflexive behavior.
This observation permits to understand what could be called the “normative versus descriptive gap”, implicit in the decision theoretic literature: because normative theories aim to prescribe, it is asked of them that they do not yield recommendations that precisely, systematically match shallow behavior. Given the numerous results in experimental psychology showing that shallow behavior is undefensible in many situations, a normative theory whose prescriptions would systematically match shallow behavior would be suspicious, to say the least. It is an immediate consequence of this minimal requirement that a normative theory cannot also be a shallow descriptive theory. 
As \citet{hausman_preference_1994} illustrate with an example of a person drinking a fatal poison while erroneously thinking that it is water, shallow behavior is not sufficient to inform normative economics.
The “normative versus descriptive gap” comes from an implicit identification of descriptive theories with shallow descriptive theories.

This gap can thus be overcome by considering reflexive descriptive theories, which is the path this article adopts.
It defines theories that are both normative and descriptive: they attempt to deliver valid prescriptions, and these prescriptions are claimed to predict reflexive behavior.

@@

\section{Previous Motivation}
It is well known that shallow preferences do not always match reflective ones.\commentYM{à quoi sert la phrase précédente ?}
Indeed, forming reflective preferences is recognised as a difficult problem in many decision problems, contrary, by definition, to shallow preferences. For example, \citet[p.\ 6]{colell_microeconomic_1995} write, when presenting the preference relation that is the basis for most of their famous book about Microeconomic theory: “Introspection quickly reveals how hard it is to evaluate alternatives that are far from the realm of common experience. It takes work and serious reflection to find out one’s own preferences.” 
In the words of \citet[p.\ 9]{bell_descriptive_1988}, “Our decision makers are not economic automatons; they make mistakes, have remorse, suffer anxieties, and cannot make up their minds”\commentYM{cette citation me semble mal venue ici, car elle ne pose pas que la question du contraste shallow/reflective, mais aussi la question de savoir s'il faudrait toujours faire un calcul économique. Par exemple avoir des remords ne me semble pas forcément une mauvaise chose} (cited by \citet[p.\ 87]{deparis_etude_2012}).
Indeed, it is to be expected that some behavior highlighted by psychologists \citep{kahneman_thinking_2012}, such as framing effects, would be rejected by the individuals themselves when pondering their decisions. \commentYM{formulation ambigue : on dirait que tu dis que les gens nieraient que les framing effects existent}

That reflective preferences are relevant to economic theory, in supplement to shallow preferences, is also illustrated by the general requirement of transitivity of preferences for economic theory, which is reasonably expected from reflective rather than from shallow preferences. As an illustration, \citet[p.\ 7]{colell_microeconomic_1995} go on writing: “the transitivity assumption can be hard to satisfy when evaluating alternatives far from common experience. As compared to the completeness property, however, it is also more fundamental in the sense that substantial portions of economic theory would not survive if economic agents could not be assumed to have transitive preferences.” \commentYM{la seconde phrase de la citation me parait superflue pour ce que tu as à dire ici. Plus généralement, la plus-value d'une citation ici me semble limitée. Plus fondamentalement, l'argument me semble incomplet. OK, les préférences shallow ont toutes les chances de n'être pas transitives. Mais pourquoi penser que les préférences réfléchies le seraient ? Il manque un argument}

Let me state how the approach proposed here to capture reflective preferences differs from those proposed in the literature, and what problems it aims at solving.

\subsection{Existing approaches for capturing reflective preferences}
%Yet, classical methods to obtain preference models in economics are either formally defined on the basis of observable data that do not describe reflective preferences, or are obtained by other means than observable preferential data and systematic reproducible formal procedures. \commentYM{je devine ce que tu veux dire, mais je ne comprends pas la dernière phrase. Il me semble important que tu t'expliques ici, car je pense que le lecteur ici ne peut pas comprendre la niche dans laquelle s'inscrit ton projet}

\subsubsection{Laundering preferences}
\label{sec:laundering}
An approach much discussed recently is the one of laundering preferences, also called preference purification \citep{bernheim_beyond_2009, rubinstein_eliciting_2011, infante_preference_2016}: having collected shallow preference data (from the field or from interviews), laundering preferences consists in transforming them, using a formal process, in such a way that they satisfy axioms generally considered as required by rationality, or keeping only preference data collected in some situations considered appropriate to determine welfare-relevant preferences. For example, \citet{bleichrodt_making_2001} transform observed preferences in order to obtain a classical Expected Utility model; \citet{bernheim_beyond_2009} propose to exclude situations where the individual seems (by objective criteria) unable to process information considered (by the theorist) as relevant information. \commentYM{les deux mécanismes évoqués sont très différents, et ont des enjeux différents. Je pense qu'il serait utile de les distinguer, et je pense qu'il faudrait en dire plus sur la nature de la transformation et du filtrage, sans quoi la discussion des ces approches est trop superficielle}

This \commentYM{what? ici on voit qu'on gagnerait à savoir concrètement quelle est la manip déployée par les auteurs ci-dessus}creates a risk of mis-representing the reflective preferences of the individual: not only is it not known whether the individual would reflectively want to adhere to these axioms \commentYM{quels axiomes ? besoin d'expliquer + en détail ce que font ces auteurs ; et tu sembles pressupposer ici qu'adéhrer aux axiomes et s'approprier une transformation ou un filtre basé sur des axiomes, c'est la même chose, ce qui ne me semble pas évident}and, if so, how she would reflectively adapt her preferences when her shallow preferences exhibit related inconsistencies; but it is not even discussed how in principle this could be known \commentYM{phrase en + super longue : arrivée à la fin, je suis perdu}. As a result, a theory postulating constraints on individual’s behavior \commentYM{je ne comprends pas de quelles contraintes sur le comportement il est question ici. Les théories des laundered preferences que je connais n'imposent aucune contrainte au COMPORTEMENT. Cela souligne encore une fois qu'il faudrait être plus prolixe sur la description des approches dont tu parles}on the basis of rational requirements is subject to the criticism addressed by \citet[p.\ 228]{dold_toward_2018} to libertarian paternalism, in that it “effectively denies the possibility that deviations from rational choice might be something to be acknowledged as reasonable”. \commentYM{phrase bien alambiquée, qui fait référence à un concept ("libertarian paternalism") que je pense abscond pour bien des gens (dont moi). Je pense que tu devrais reformuler, donner l'argument plus directement, et si tu y tiens mettre entre parenthèse que cet arguement est utilisé dans un autre débat par x et y quand ils parlent de LP}
That this is an important methodological problem is sometimes recognized by authors adopting \commentYM{endorsing}this approach \commentYM{je pense qu'en anglais il est + idiomatique d'éviter ce genre de tournure : plutôt dire "Authors... sometimes recognize that..."}. \Citet{bleichrodt_making_2001}, for example, state that laundering require controversial assumptions (p.\ 1500), and recommend it only as a last resort, recommending discussing with the decision-maker whenever possible to elicit her reflective preferences (p.\ 1499). Note that nudging \citep{thaler_nudge_2009} creates a similar problem: we have to guess what the individual would like if he thought hard about the decision problem, was well-informed, and actively worked against the psychological features he himself does not consider adequate for good decision making.

@@

\subsubsection{Informal deliberation}
Another path to obtain reflectful \commentYM{reflective ?} preferences has been followed: the one of deliberation, an umbrella term by which I designate processes, either, involving discussions between an analyst and a decision-maker while querying her preferences in view of building a formal model of it \commentYM{them ?} \citep{raiffa_back_1985, keeney_decisions_1993, roy_multicriteria_1996, belton_multiple_2002}, or involving group discussions, with presentations from experts to an audience and internal deliberation phases \citep{fishkin_when_2011}, aiming to obtain well-informed preferences. \commentYM{il me semble nécessaire ici de mentionner aussi tous les dispositifs participatifs et de concertation employés en routine dans les prises de décisions relatives aux politiques publiques et leur mise en oeuvre}

A major problem of these techniques is that they are not easily reproducible \commentYM{il faudrait expliquer en quel sens}, and that the resulting preferences elicited from the subjects might reflect properties of the personalities of the intervening experts or group-dynamic effects rather than true reflective preferences of the individuals \commentOCf{Yves, as-tu une référence pour moi ?}\commentYM{on a mis des références pertinentes pour les différents aspects dan le papier soumis à Ecol Econ. Ces références ne te conviennent pas ?}. This problem is exacerbated \commentYM{compounded} by the fact that psychologists have shown that apparent coherence can appear, that is, preferences that do not violate axioms usually considered as representing rationality, although some framing effect may be at play that result in the apparently coherent preferences to arguably not represent reflective preferences \citep{ariely_predictably_2010} \commentYM{phrase beaucoup trop longue pour être compréhensible, qui fait par ailleurs référence au concept de rationalité sans le définir, ce qui est acceptable en début d'intro, mais me semble problématique à ce stade dans un papier qui cherche à éclairer les débats flous par des clarifications formelles et conceptuelles}. It also remains to be clarified how to check on a principled, systematic basis, whether all relevant arguments have been presented faithfully. Comfronted with (possibly self-designated \commentYM{ceci est un tout autre problème, dont la mention ici brouille le propos}) experts with enough assertiveness, a decision-maker can be convinced to follow principles of decision-making that several experts in the field of decision-theory \commentYM{would} consider incorrect. As an illustration, after discussing the weaknesses of the \ac{AHP}, \citet[p.\ 53]{howard_foundations_2007} write \commentYM{s}: “Why, then, do inferior processes find favor with decision makers? The answer is that they do not force you to think very hard or to think in new ways.” \commentYM{je ne vois pas en quoi ceci est censé illustrer le propos précédent (ou alors je n'ai rien compris à la citation). J'ai l'impression que la citation dit que, parfois, il y a des forces psychologiques qui s'opposent à l'adoption d'un processus de décision, quand bien même celui-ci aurait des qualités indéniables}
Beyond the controversial \ac{AHP}, famous decision analysts have voiced disagreements with normative theories of decision making \citep{ellsberg_risk_1961, allais_so-called_1979} that are however generally applied in the field. \commentYM{il me semble ici intenable de faire l'économie de définir ce que tu appelles "normative theories"} 
\Citet[p.\ 669]{ellsberg_risk_1961} concluded his influential article as follows: “[In some situations, for some people,] the Bayesian or Savage approach gives [...] bad advice. [Some people] act in conflict with the axioms deliberately, without apology, because it seems to them the sensible way to behave. Are they clearly mistaken?” \commentYM{d'une manière générale, dans toute la section 1, l'avalanche de citation me semble nuire au propos. Cela induit des ruptures dans le style, et force le lecteur à se pencher sur des questions d'interprétation des citations qui l'empêchent de se focaliser sur ton propos à toi}
This rethorical question illustrates the need of systematic procedures able to determine whether decision-makers, when confronted with the relevant arguments, accept the soundness of the principles on whose basis the recommendation rests \commentYM{the principles that the recommendation rests on}(this has been somewhat explored, but not systematically, for current lack \commentYM{for lack} of a systematic way of confronting decision makers to arguments \citep{slovic_who_1974, stanovich_discrepancies_1999}).

As a consequence of these techniques being not \commentYM{not being}completely formalised and reproducible, each application requires manual intervention \commentYM{quelles sont ces "interventions manuelles" ? à expliquer}of (very skilled) experts. Eliciting reflective preferences therefore requires enormous resources. \Citet[p.\ 572]{pinto_normative_2012} observe that “in many of those surveys that governments use in order to inform social policy there are not enough resources (time and money) in order to reach the stage where preferences are not labile”. There is, to reduce this cost, a need for theories that can be iteratively improved upon and yield reusable knowledge about reflective preferences. \commentYM{Cette idée de limiter le coût est intéressante, et je me dis qu'elle peut faire mouche avec certains lecteurs. Mais en l'état ça ne fonctionne pas car tu ne expliques pas quels sont les coûts impliqués dans les approaches "manuelles"}

Different analysts, using different methods and different axioms as their favorite conception of rational decision-making, will obtain possibly contradictory conclusions, “elicited” from the decision-makers who hire them. Some analysts admit this possibility and do not consider it to be a problem. \commentYM{il me semble que tu soulèves là encore un nouveau problème : comme la notion de rationalité n'est pas claire, il y a débat sur les axiomes qu'on peut vouloir lui associer, et du coup différentes approches, choisissant différents axiomes, vont éliciter des préférences différentes. Je te rejoins sur le fait que c'est un problème important, mais il est différent de ce qui a été discuté juste avant, et tu ne l'explicites pas. Il me semble qu'il faudrait bien distinguer les problèmes, et bien les formuler, et il faudrait que le plan soit plus clair et lisible. Là, en lisant, je ne sais pas si tu considères ce que tu abordes dans ce § comme un nouveau problème, ou comme un aspect ou une implication, du problème exploré précédemment}
For example, \citet[p.\ 214]{roy_comparison_1995} write: “In our view, these inevitable disagreements do not imply that decision-aid is useless but simply that a single problem may have several valid responses. Given that two different decision-aid models cannot be implemented in the same decision process, the decision-maker must be conscious of the qualitative choices implied by the different models – often conveying the analysts’ own ethical choices – before coming to personal conclusions on the choice to be made. In this domain, the many different approaches reflect in our view the complexity of the researcher's task much more than a scientific weakness.”\commentYM{ici encore, je fais une overdose de citations. Et je ne trouve pas celle-ci particulièrement éclairante pour ton propos. Certes il y a un lien entre ce que tu dis et ce que dit cette citation, mais elle n'est pas vraiment une illustration de ce que tu dis. Du coup elle brouille le message}
Although being an important insight about the (likely vain) hope of determining a single correct recommendation in many complex decision situations (about which more below), this answer \commentYM{quelle answer ? à quelle question ? je suis perdu}does not address the problem that some principles of decision making may be wrong guides in some situations, possibly including some principles that are commonly applied in the field, hence the need of principled ways of discriminating them. \commentYM{est-ce vraiment ton projet ? as-tu l'intention de trouver la meilleure des axiomatiques de décision ? si c'est ton projet, il faut le dire clairement. Ce que je comprennais jusqu'ici était que tu cherchais un moyen de connaître les "preferences that represent informed, pondered decisions, that would be taken after due consideration by the individual", ce qui me semble être un tout autre projet}

>>>>

\section{Species of theories}
\label{sec:approaches}
\subsection{Overview}
For this article, a \emph{normative decision theory} is a theory that, when applied in the situations it self-describes as applicable to, claims to yield valid prescriptions. I consider a prescription in the wide sense: recommending against some behavior, or recommending conditionally, also counts as prescribing.

A \emph{descriptive decision theory} is a theory that, when correctly applied in the situations it self-describes as applicable to, claims to predict the decision behavior of a given set of individuals. For this article, a claim is a prediction only when it is in principle possible to show that it is false using empirical data.

@@

These definitions avoid to define the concept of theory itself, for defining precisely such a fundamental concept raises serious difficulties. Is Prospect theory in general a theory, or only some of its variants, sufficiently precisely defined? Can a theory be extremely specific, or is some level of generality required to deserve being called a theory? Must a theory consist in a non-trivial assembly of laws, some deducible from others? It is not necessary for the purpose of the present discussion to take position on these issues, rather, the focus is on the distinction between normative and descriptive approaches. An intuitive understanding of the concept of theory will suffice.
Similarly, I sidestep insuperable difficulties pertaining to capturing in a short definition the concept of a descriptive theory in general: some descriptive theories in a more general sense may propose to describe without predicting, for example (historical decisions or trends come to mind); such endeavor simply does not enter the scope of this discussion. For similar reasons, I do not claim that the term normative is unanimously used in the literature as defined here. This definition encompasses theories used in approaches that do not necessarily self-describe as normative \citep{bell_decision_1988, meinard_rationality_2019}; conversely, some theories considered as normative in the literature, such as some social choice theories, may not fall into the definition proposed here, depending on how they are interpreted.

I say that a normative theory is also a descriptive theory when the prescriptions of the normative theory are also claimed to predict the decision behavior of a given set of individuals.

Another contrast can now be introduced: the contrast between shallow and reflective behavior. \emph{Shallow behavior} is defined as the everyday, usual behavior of an individual. 
This is the behavior whose observation defines what \citet[p.\ 16]{von_neumann_theory_2004} describe as the “intuitive” preferences of an individual, those originating from an “immediate sensation” (cited in \citet{fishburn_retrospective_1989}).
\emph{Reflective behavior}, by contrast, is defined as the behavior that enacts informed, pondered decisions, those that would be taken after due consideration.
% an individual has consciously adopted after having pondered everything that is possibly relevant to the decision. 
%This pair of definitions is, again, vague and unsatisfactory for a deep philosophical discussion, but hopefully sufficient for the present one. 
%What is meant by “pondering about everything that is possibly relevant to the decision” will be defined more precisely later; an intuitive understanding of this phrase will do at this stage.
This distinction draws on our capacity, as human beings, to think about our shallow preferences and consider some of them as mistaken \citep{frankfurt_freedom_1971}. 
A famous example of such a situation is the one of \citet[pp.\ 101--103]{savage_foundations_1972} having realized, after due consideration, that his initial declared preference in a decision problem presented by \citet{allais_so-called_1979} was mistaken.

This contrast raises a new distinction, which appears by simple combination of the just given definitions: a \emph{shallow descriptive theory} is a descriptive theory that claims to predict the shallow behavior of individuals; a \emph{reflective descriptive theory} is one that claims to predict the reflective behavior of individuals.

The descriptive theories proposed in decision theory, in the sense defined here, are exclusively shallow descriptive theories, to the best of my knowledge.
%This need not be so. Indeed, this article introduces descriptive theories about reflective behavior.
This observation permits to understand what could be called the “normative versus descriptive gap”, implicit in the decision theoretic literature: because normative theories aim to prescribe, they are required to not yield recommendations that precisely, systematically match shallow behavior. Given the numerous results in experimental psychology showing that shallow behavior is undefensible in many situations, a normative theory whose prescriptions would systematically match shallow behavior would be suspicious, to say the least. It would also be of low practical interest.
It is an immediate consequence of this minimal requirement that a normative theory cannot also be a shallow descriptive theory. 
In other words, as \citet{hausman_preference_1994} illustrate with an example of a person drinking a fatal poison while erroneously thinking that it is water, shallow behavior is not sufficient to inform normative economics.
The “normative versus descriptive gap” comes from an implicit identification of descriptive theories with shallow descriptive theories.

This gap can thus be overcome by considering reflective descriptive theories, instead of shallow descriptive theories. This is the path this article adopts.
It defines theories that are both normative and descriptive: they attempt to deliver valid prescriptions, and these prescriptions are claimed to predict reflective behavior.

@@

Let me state how the approach proposed here to capture reflective preferences differs from those proposed in the literature.

\subsection{Existing approaches for modeling reflective preferences}
The review of existing approaches aiming at obtaining a model of the reflective preferences of an individual distinguishes two kinds of approaches. \Cref{sec:purification} mentions formal techniques, that can be applied by following an explicit procedure, but that do not consider the reaction of individuals to arguments; while \cref{sec:informal} reviews studies that involve informal discussions with individuals, which do consider arguments, but require involving experts. By contrast, this article presents theories that can be confronted to empirical data in a formal, systematic way, and that consider the reactions of individuals to arguments.

Before delving into these two categories of approaches, let me mention a few others that, though related to the topic discussed here, fall outside the scope of this article.

Some authors have proposed to base normative theories on observable aspects of nature and individuals; for example, \citet{kahneman_back_1997} propose to measure “experienced utility”, considered as a hedonic quality. 
To the best of my knowledge, these proposals never aim at observing reflective preferences as considered here and thus fall outside the scope of this article.

\subsubsection{Preference purification}
\label{sec:purification}
An approach much discussed recently is the one of preference purification \citep{bernheim_beyond_2009, rubinstein_eliciting_2011, infante_preference_2016}.
Having collected shallow preference data (from the field or from interviews), purifying preferences consists in transforming them, using a formal process, in such a way that they satisfy axioms generally considered as required by rationality, or keeping only preference data collected in some situations considered appropriate to determine welfare-relevant preferences. 
For example, \citet{bleichrodt_making_2001} transform observed preferences in order to obtain a classical Expected Utility model; \citet[Section VII]{bernheim_beyond_2009} propose to exclude situations where the individual seems (by objective criteria) unable to process information considered (by the theorist) as relevant information.%; \citet{rubinstein_eliciting_2011} study the possible “rational” preference relations that are coherent with the observed data when assuming some distortion process relating (shallow, observed) and (reflective, unobserved) behavior. 
\commentYM{les deux mécanismes évoqués sont très différents, et ont des enjeux différents. Je pense qu'il serait utile de les distinguer, et je pense qu'il faudrait en dire plus sur la nature de la transformation et du filtrage, sans quoi la discussion des ces approches est trop superficielle}
\commentOC{Je ne vois pas comment faire ça sans en faire une digression énorme. Il faudrait introduire des notations, des définitions, … Je vois à peu près comment les présenter sous un cadre unifié et les comparer axiomatiquement aux théories du DJ, mais cela devrait faire l’objet d’un autre article, il me semble.}

These proposals leave unspecified how to confirm or disconfirm them empirically, thus, in particular, how to empirically investigate disagreements between theories: if two theories, on the basis of the same observed data, postulate that they correspond to different purified preferences, the disagreement may at best be arbitrated using ad-hoc, case by case criteria to assess for their relative plausibility. By contrast, the present proposal indicates a clear, systematic empirical way to abitrate disagreements.
%These proposals deserve a detailed comparison to the current one, but this requires a treatment that cannot be given in this article. Let me simply highlight one major difference. which can be done with a high level description of these approaches using a variant of the framework that \citet{rubinstein_eliciting_2011} describe.
%Consider a set of possible observations $S$, describing shallow behavior, and a set of possible “rational” preference relations $R$. The set $S$ may contain observations of choice instances, for example, given the bundle $B$, the individual chose $b \in B$; or may contain all possible linear orders over some set of alternatives. The set $R$ contains all relations that satisfy some set of axioms of rationality, representing what the modeler considers as an acceptably purified relation. \Citet{bleichrodt_making_2001}, for example, use for $R$ the preference relations satisfying \ac{SEU}.

%The purification process amounts to define a function $f: \powerset{S} \rightarrow \powerset{R}$, where $\powerset{A}$ represents the set of subsets of a given set $A$. The function $f$ relates a set of observations to their possible purified preference relations. The argument of the function represents observations obtained using different frames. se could be preference orders 
%This \commentYM{what? ici on voit qu'on gagnerait à savoir concrètement quelle est la manip déployée par les auteurs ci-dessus}creates a risk of mis-representing the reflective preferences of the individual: not only is it not known whether the individual would reflectively want to adhere to these axioms 
%\commentYM{quels axiomes ? besoin d'expliquer + en détail ce que font ces auteurs ; et tu sembles pressupposer ici qu'adéhrer aux axiomes et s'approprier une transformation ou un filtre basé sur des axiomes, c'est la même chose, ce qui ne me semble pas évident}
%and, if so, how she would reflectively adapt her preferences when her shallow preferences exhibit related inconsistencies; but it is not even discussed how in principle this could be known \commentYM{phrase en + super longue : arrivée à la fin, je suis perdu}. As a result, a theory postulating constraints on individual’s behavior \commentYM{je ne comprends pas de quelles contraintes sur le comportement il est question ici. Les théories des laundered preferences que je connais n'imposent aucune contrainte au COMPORTEMENT. Cela souligne encore une fois qu'il faudrait être plus prolixe sur la description des approches dont tu parles}on the basis of rational requirements is subject to the criticism addressed by \citet[p.\ 228]{dold_toward_2018} to libertarian paternalism, in that it “effectively denies the possibility that deviations from rational choice might be something to be acknowledged as reasonable”. \commentYM{phrase bien alambiquée, qui fait référence à un concept ("libertarian paternalism") que je pense abscond pour bien des gens (dont moi). Je pense que tu devrais reformuler, donner l'argument plus directement, et si tu y tiens mettre entre parenthèse que cet arguement est utilisé dans un autre débat par x et y quand ils parlent de LP}
In other words, it is left unclear how, from a given observation of shallow behavior, one may obtain empirically information about reflective (or purified, in the terminology of this literature) preferences, or about the axioms they should satisfy. Authors endorsing this approach sometimes recognize that this is an important methodological problem. \Citet{bleichrodt_making_2001}, for example, state that their proposal require controversial assumptions (p.\ 1500), and recommend it only as a last resort, recommending discussing with the decision-maker whenever possible to elicit her reflective preferences (p.\ 1499). Note that nudging \citep{thaler_nudge_2009} creates a similar problem: we have to guess what the individual would like if he thought hard about the decision problem, was well-informed, and actively worked against the psychological features he himself does not consider adequate for good decision making.

@@

\subsubsection{Informal deliberation}
\label{sec:informal}
Another path to obtain reflective preferences has been followed: the one of deliberation, an umbrella term by which I designate processes, either, involving discussions between an analyst and a decision-maker while querying her preferences in view of building a formal model of them \citep{raiffa_back_1985, keeney_decisions_1993, roy_multicriteria_1996, belton_multiple_2002}, or involving group discussions, with presentations from experts to an audience and internal deliberation phases \citep{fishkin_when_2011, chappell_deliberative_2012}, aiming to obtain well-informed preferences. 
The term as used here also includes the informal use of debiasing techniques: techniques that aim at encouraging individuals to think more about their decision problem in order to reduce occurrence of behavioral “mistakes”. Examples include suggesting the individuals to consider alternative points of view, or asking them to give reasons for their choices (an insightful article by \citet{mitchell_libertarian_2004} cite many such studies). 
\commentYM{il me semble nécessaire ici de mentionner aussi tous les dispositifs participatifs et de concertation employés en routine dans les prises de décisions relatives aux politiques publiques et leur mise en oeuvre}
\commentOC{Je m’interroge. Ça me semble de pus en plus éloigné du scope, car je m’intéresse aux préférences d’un individu donné, hors questions d’agrégation.}

A major problem of these techniques is that they are not easily reproducible \commentYM{il faudrait expliquer en quel sens}: the resulting preferences elicited from the subjects might reflect properties of the personalities of the intervening experts or group-dynamic effects rather than true reflective preferences of the individuals.

This problem is compounded by the phenomenon of arbitrary coherence, exhibited by psychologists \citep{ariely_predictably_2010}.  This term denotes situations where individuals form an apparently coherent system of preference that depends on the path taken to elicit it. 
\commentYM{phrase beaucoup trop longue pour être compréhensible, qui fait par ailleurs référence au concept de rationalité sans le définir, ce qui est acceptable en début d'intro, mais me semble problématique à ce stade dans un papier qui cherche à éclairer les débats flous par des clarifications formelles et conceptuelles}.

It also remains to be clarified how to check on a principled, systematic basis, whether all relevant arguments have been presented faithfully. Comfronted with 
%(possibly self-designated \commentYM{ceci est un tout autre problème, dont la mention ici brouille le propos}) 
experts with enough assertiveness, a decision-maker can be convinced to follow principles of decision-making that several experts in the field of decision-theory would consider incorrect. 
%As an illustration, after discussing the weaknesses of the \ac{AHP}, \citet[p.\ 53]{howard_foundations_2007} write \commentYM{s}: “Why, then, do inferior processes find favor with decision makers? The answer is that they do not force you to think very hard or to think in new ways.” \commentYM{je ne vois pas en quoi ceci est censé illustrer le propos précédent (ou alors je n'ai rien compris à la citation). J'ai l'impression que la citation dit que, parfois, il y a des forces psychologiques qui s'opposent à l'adoption d'un processus de décision, quand bien même celui-ci aurait des qualités indéniables}
Beyond the controversial \ac{AHP} mentioned in the introduction, famous decision analysts have voiced disagreements with normative theories of decision making \citep{ellsberg_risk_1961, allais_so-called_1979} that are however generally applied in the field. \commentYM{il me semble ici intenable de faire l'économie de définir ce que tu appelles "normative theories"} 
\Citet[p.\ 669]{ellsberg_risk_1961} concluded his influential article as follows: “[In some situations, for some people,] the Bayesian or Savage approach gives [...] bad advice. [Some people] act in conflict with the axioms deliberately, without apology, because it seems to them the sensible way to behave. Are they clearly mistaken?” \commentYM{d'une manière générale, dans toute la section 1, l'avalanche de citation me semble nuire au propos. Cela induit des ruptures dans le style, et force le lecteur à se pencher sur des questions d'interprétation des citations qui l'empêchent de se focaliser sur ton propos à toi}
This rethorical question illustrates the need of systematic procedures able to determine whether decision-makers, when confronted with the relevant arguments, accept the soundness of the principles that the recommendation rests on (this has been somewhat explored, but not systematically, for lack of a systematic way of confronting decision makers to arguments \citep{slovic_who_1974, stanovich_discrepancies_1999}).

Another way to view this problem is that different analysts, using different methods and different axioms as their favorite conception of rational decision-making, will obtain possibly contradictory conclusions, “elicited” from the decision-makers who hire them. Some analysts admit this possibility and do not consider it to be a problem. \commentYM{il me semble que tu soulèves là encore un nouveau problème : comme la notion de rationalité n'est pas claire, il y a débat sur les axiomes qu'on peut vouloir lui associer, et du coup différentes approches, choisissant différents axiomes, vont éliciter des préférences différentes. Je te rejoins sur le fait que c'est un problème important, mais il est différent de ce qui a été discuté juste avant, et tu ne l'explicites pas. Il me semble qu'il faudrait bien distinguer les problèmes, et bien les formuler, et il faudrait que le plan soit plus clair et lisible. Là, en lisant, je ne sais pas si tu considères ce que tu abordes dans ce § comme un nouveau problème, ou comme un aspect ou une implication, du problème exploré précédemment}
\commentOCf{C’est discutable si c’est un autre problème ou un peu le même problème ou carrément le même problème vu autrement. Du point de vue logique, toutes les propositions fausses sont équivalentes. En l’absence d’une analyse comparative approfondie de toutes ces approches et de leurs propriétés, je ne vois pas à ce stade comment faire beaucoup plus clair. Mais je serais heureux d’en discuter si tu vois une manière de patcher le problème raisonnablement facilement.}
For example, \citet[p.\ 214]{roy_comparison_1995} write: “In our view, these inevitable disagreements do not imply that decision-aid is useless but simply that a single problem may have several valid responses. Given that two different decision-aid models cannot be implemented in the same decision process, the decision-maker must be conscious of the qualitative choices implied by the different models – often conveying the analysts’ own ethical choices – before coming to personal conclusions on the choice to be made. In this domain, the many different approaches reflect in our view the complexity of the researcher's task much more than a scientific weakness.”\commentYM{ici encore, je fais une overdose de citations. Et je ne trouve pas celle-ci particulièrement éclairante pour ton propos. Certes il y a un lien entre ce que tu dis et ce que dit cette citation, mais elle n'est pas vraiment une illustration de ce que tu dis. Du coup elle brouille le message}
\commentOCf{J’ai un peu réduit les citations, j’espère que ça va mieux. Concernant celle-ci en particulier, elle me semble très utile. Je trouve stupéfiant qu’on puisse trouver ces propos dans la littérature, écrits noir sur blanc. Si je ne les cite pas, un lecteur pourrait (ou devrait) ne pas me croire quand j’affirme que des gens trouvent que ce n’est pas grave que deux études soi-disant scientifiques arrivent à des conclusions contradictoires. La citation étendue permet de comprendre, je crois, leur point de vue, et de se faire une idée de ce qu’ils veulent dire par là.}
Although being an important insight about the (likely vain) hope of determining a single correct recommendation in many complex decision situations (about which more below), this insight does not address the problem that some principles of decision making may be wrong guides in some situations, possibly including some principles that are commonly applied in the field, hence the need of principled ways of discriminating them. \commentYM{est-ce vraiment ton projet ? as-tu l'intention de trouver la meilleure des axiomatiques de décision ? si c'est ton projet, il faut le dire clairement. Ce que je comprennais jusqu'ici était que tu cherchais un moyen de connaître les "preferences that represent informed, pondered decisions, that would be taken after due consideration by the individual", ce qui me semble être un tout autre projet}
\commentOCf{Je propose de trancher entre approches normatives en utilisant des données empiriques, plutôt que simplement la force de conviction des axiomes (qui parlent différemment à différentes personnes et peuvent conserver les désaccords intacts). Je ne sais pas si c’est plus clair maintenant.}
